Fail Faster

### 2025-03-17 paper/paper.tex
wibbles

Just how much runtime overhead does this monadic abstraction introduce? To
illustrate, let's return to our running example of a constrained pair of
integers, written in both \bq{} and ScalaCheck.
Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
The first version (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) is written with monadic generator combinators.
\jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
The second version (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
is just the first version, with (a) all generator combinator definitions inlined, and (b)
beta-redexes reduced.

@@

def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
}
\caption{Int Pair Generators in \bq{} and ScalaCheck}
\label{fig:overhead-explanation-code}
\end{figure}

The performance difference the two versions of this generator (Figure~\ref{fig:overhead-explanation-perf}) is stark. In both
languages, the inlined version takes \textbf{half} as much time to generate a
single pair of \texttt{int}s.

\begin{figure}
\end{lstlisting}
  \begin{tabular}{lll}
  Library    & Generator          & Time per Generation (ns) \\
  \bq{}         & \texttt{int\_pair}          & 70                       \\
  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
  ScalaCheck & \texttt{intPair}            & 458                      \\
  ScalaCheck & \texttt{intPairInlined}     & 266                     
  \end{tabular}
\label{fig:overhead-explanation-perf}
\end{figure}

% ┌───────┬──────────┬─────────┬────────────┐
% │ Name  │ Time/Run │ mWd/Run │ Percentage │
% ├───────┼──────────┼─────────┼────────────┤
% │ bq    │  70.47ns │  78.00w │    100.00% │
% │ fused │  35.86ns │  78.00w │     50.89% │
% └───────┴──────────┴─────────┴────────────┘

@@

\jwc{
The problem is plain: while the monadic abstraction that PBT libraries like base quickcheck provide are indespensible for writing idiomatic generators, the performance overhead of using them is dramatic.
Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
}

@@

\subsection{Combinator Abstraction Overhead}

@@

This type, along with basic monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The monadic interface is given by a return and bind, as usual.
\texttt{Return} is the constant generator, but this time it 
runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
returns the generator which always generates that value.
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation. We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}

>>>>

Just how much runtime overhead does this monadic abstraction introduce? To
illustrate, let's return to our running example of a constrained pair of
integers, written in both \bq{} and ScalaCheck.
Figure~\ref{fig:overhead-explanation-code} shows two versions of the generator, written in both languages.
The first versions (\texttt{int\_pair} in \bq{} and \texttt{intPair} in ScalaCheck) are written with the monadic generator combinators from their respective libraries.
\jwc{Maybe we can avoid rewriting these if we already showed them earlier? Also maybe we duplicate the tutorial discussion of monadic generator libraries to include Scala?}
The second versions (\texttt{int\_pair\_inlined} and \texttt{intPairInlined})
are semantically identical to the first, but have been rewritten
by (1) inlining all generator combinator definitions, and then (2)
repeatedly reducing simplifiable terms like \lstinline{(fun x -> e) e'} --- where an anonymous function is defined and then immediatly called~\footnote{This is usually known as a ``$\beta$-redex''.
Note that when reducing, we do not reduce to \lstinline{e[e'/x]}, as this does not preserve the order of effects.
}
--- to \lstinline{let x = e' in e}.

@@

def intPairInlined : (Gen.Parameters, Seed) => (Option[(Long,Long)],Seed) = {
 (p,seed) =>
  val (x,seed2) = chLng(0,1000)(p,seed)
  x match {
    case None => (None,seed2)
    case Some(x) =>
      val (y,seed3) = chLng(0,x)(p,seed2)
      y match {
        case None => (None,seed)
        case Some(y) => (Some(x,y),seed3)
      }
    }
}
\end{lstlisting}
\caption{Int Pair Generators in \bq{} and ScalaCheck}
\label{fig:overhead-explanation-code}
\end{figure}

The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes (on average) \textbf{half} as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
\jwc{more precise ref}) show an even more dramatic performance boost for
inlining in this manner.

% Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
% These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
% The

\begin{figure}
  \begin{tabular}{lll}
  Library    & Generator          & Average Time per Generation (ns) \\
  \bq{}         & \texttt{int\_pair}          & 70                       \\
  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
  ScalaCheck & \texttt{intPair}            & 458                      \\
  ScalaCheck & \texttt{intPairInlined}     & 266                     
  \end{tabular}
\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn)}
\label{fig:overhead-explanation-perf}
\end{figure}

@@

% \jwc{
% 
% }

@@

\subsubsection{Overhead of Other Combinators}

@@

This type, along with basic staged monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
\texttt{Return} is the constant generator, but this time it 
runs at compile time. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it
returns the generator which always generates that value.
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one \footnote{Readers familiar with OCaml may notice that \texttt{return : 'a code -> 'a Gen.t} and \texttt{bind : 'a Gen.t -> ('a code -> 'b Gen.t) -> 'b Gen.t}
do not have the correct types for a monad instance, preventing us from using \texttt{let\%bind} notation. We rectify this issue in Section~\ref{subsection:codecps} by importing some clever ideas from the staging literature.}.
Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
